On the spatial aggregation of condition metrics for ecosystem accounting

### 2025-12-01 manuscript/manuscript.qmd
change title

---
title: On the spatial aggregation of ecosystem condition indicators
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Hanno Sandvik
    email: hanno.sandvik@nina.no
    orchid: https://orcid.org/0000-0002-5889-1606
    affiliations:
      - ref: nina
  - name: Bálint Czúcz
    email: balint.czucz@nina.no
    affiliations:
      - ref: nina
  - name: Chloé R. Nater
    email: chloe.nater@nina.no
    affiliations:
      - ref: nina
abstract: |
  In face of the ongoing nature crisis, the international community is setting targets and deciding on actions to combat the current biodiversity crises. For this to be effective they need tools to accurately describe the current situation and to monitor trends in ecosystems over time. Ecosystem condition accounts (ECA) is one such tools that use variables and indicators to describe key ecosystem characteristics, reflecting their condition and deviations from a reference condition. Because the purpose is to inform decisions at relatively high political levels, these metrics are often spatially aggregated to represent larger areas, such as countries. However, spatial aggregation of information has the potential to alter the descriptive and normative interpretations one can make from these metrics. For example, aggregation displacement causes the information held in variables and indicators to diverge when these are aggregated spatially. This process is influenced also by the order of steps involved in normalising and aggregating variables, i.e. the aggregation pathway. Although aggregation displacement and the type of aggregation pathway chosen for the indicator clearly impact both the indicator values and their interpretation, there are no clear guidelines or deliberation on these topics in the SEEA EA standard for ecosystem accounting. This paper outlines the consequences of different aggregation pathways, emphasising their impact on the credibility of ECAs, and how these are interpreted by users. We introduce a standardised terminology for aggregation pathways specific to ecosystem condition indicators following the SEEA EA standard and provide recommendations for selecting appropriate pathways in various contexts. Our discussion of this topic is aimed at raising the general awareness of spatial aggregation issues and to guide indicator developers in choosing and reporting spatial aggregation methods.

@@

In our experience, the choice of which area units to use when doing this step-wise aggregation is often not reflected much upon.
In some cases the areal units are even decided politically, for example to administrative units, with little ecological justification for this choice.
However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to as the **modifiable area unit problem (MAUP)**.
The MAUP can lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
This is unfortunate, as it is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
One option to reduce this discrepancy and arbitrariness is to use ecologically motivated areal units in the step-wise aggregation such as homogeneous ecosystem areas [@vallecillo2022] or homogeneous impact areas [@kolstad_making_2024].

@@

Pathway 4 illustrates the aggregation of a variable, with no normalisation.
This is the pathway commonly used for variable accounts in the SEE EA. For this, and all other aggregation pathways, the picture can be made more complicated by adding additional aggregation levels (i.e. a step-wise aggregation, @sec-comp)

@@

Pathways 8 and 9 describe alternative ways to create indices at different spatial scales, either by spatially aggregating the indices (pathway 8), or by aggregating the indicators separately and recalculating the indices at different spatial scales (pathway 9).
If perfectly commutative then the two approaches will yield the same result.
However, if the indicators that make up the indices use different area weights as part of their spatial aggregation (**Example 4**), then pathway 9 is the only one that gives the same index values across nested accounts (@sec-comp).

@@

ECAs should be accompanied by detailed indicator documentation where a precise description of the steps in the aggregation process is presented.
Verbose descriptions can be supplemented or replaced by a standardised short-hand notation that we present in the following: We suggest using the terms as described in @fig-term and elsewhere in the paper: scale, truncate (abbreviated to *trunc.*), transform (*trans.*), spatial aggregation (*sp. agg.*), thematic aggregation (*th. agg.*) and spatial statistics (*sp. stat.*).
These terms can be abbreviated as shown in the parentheses, and placed together into a single string, where operations performed simultaneously are enclosed in square brackets.
For example, pathway 1 in @fig-pathways can be annotated as $scale - trunc. - trans. - sp. agg.$.
If the three normalisation steps are performed simultaneously this can be written as $[scale - trunc. - trans.] - sp. agg.$.
If there are more aggregation steps, then the spatial level at each step should be included in the description, e.g. for pathway 3: $sp. agg. (municipalities) - scale - sp. agg. (country)$.
"Municipality" and "country" could here be included as footnotes.
Note also that in the last example, normalisation only involved a scaling step.
The description of indicator pathways can be made even easier by referencing specific aggregation pathways by number, as we have done in this paper using @fig-pathways in this paper.
Methods used to change either the cover or the extent of the indicator data (see @sec-term) can also be included in the shorthand notation scheme.
The three operations we conceive as possible options are extrapolation (extr.), interpolation (inter.), and (sub)sampling.
We use the term spatial projection (projection for short) to encompass both extra- and interpolation.
Thus, pathway 7 in Fig.
3 can be written as $project - scale - sample - sp.agg. - trunc$.
The type of model used for the spatial projection can be added as a footnote.

@@

It may be premature, or not even possible, to prescribe an aggregation pathway to be used for all ECAs, or even a type of aggregation pathway to be used given the data at hand.
However, internal consistency within assessments should be possible, and this can make it easier to communicate and interpret the information conveyed through the indicator values.
This can be achieved by for example always scaling early (see recommendation number 2 above) and using the same aggregation levels (same spatial units) and same approach to step-wise aggregation across levels (@sec-comp), for all indicators.

@@

...

>>>>

---
title: On the spatial aggregation of condition metrics for ecosystem accounting
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Hanno Sandvik
    email: hanno.sandvik@nina.no
    orchid: https://orcid.org/0000-0002-5889-1606
    affiliations:
      - ref: nina
  - name: Bálint Czúcz
    email: balint.czucz@nina.no
    affiliations:
      - ref: nina
  - name: Chloé R. Nater
    email: chloe.nater@nina.no
    affiliations:
      - ref: nina
abstract: |
  In face of the ongoing nature crisis, the international community is setting targets and deciding on actions to combat the current biodiversity crises. For this to be effective they need tools to accurately describe the current situation and to monitor trends in ecosystems over time. Ecosystem condition accounts (ECA) is one such tools that use variables and indicators to describe key ecosystem characteristics, reflecting their condition and deviations from a reference condition. Because the purpose is to inform decisions at relatively high political levels, these metrics are often spatially aggregated to represent larger areas, such as countries. However, spatial aggregation of information has the potential to alter the descriptive and normative interpretations one can make from these metrics. For example, aggregation displacement causes the information held in variables and indicators to diverge when these are aggregated spatially. This process is influenced also by the order of steps involved in normalising and aggregating variables, i.e. the aggregation pathway. Although aggregation displacement and the type of aggregation pathway chosen for the indicator clearly impact both the indicator values and their interpretation, there are no clear guidelines or deliberation on these topics in the SEEA EA standard for ecosystem accounting. This paper outlines the consequences of different aggregation pathways, emphasising their impact on the credibility of ECAs, and how these are interpreted by users. We introduce a standardised terminology for aggregation pathways specific to ecosystem condition indicators following the SEEA EA standard and provide recommendations for selecting appropriate pathways in various contexts. Our discussion of this topic is aimed at raising the general awareness of spatial aggregation issues and to guide indicator developers in choosing and reporting spatial aggregation methods.

@@

In our experience, the choice of which area units to use when doing this step-wise aggregation is often not reflected much upon.
In some cases the areal units are even decided politically, for example to administrative units, with little ecological justification for this choice.
However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to as the **modifiable area unit problem (MAUP)**.
The MAUP can lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust.
This is unfortunate, as it is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
One option to reduce this discrepancy and arbitrariness is to use ecologically motivated areal units in the step-wise aggregation such as homogeneous ecosystem areas [@vallecillo2022] or homogeneous impact areas [@kolstad_making_2024].

@@

Pathway 4 illustrates the aggregation of a variable, with no normalisation.
This is the pathway commonly used for variable accounts in the SEE EA. For this, and all other aggregation pathways, the picture can be made more complicated by adding additional aggregation levels (i.e. a step-wise aggregation).

@@

Pathways 8 and 9 describe alternative ways to create indices at different spatial scales, either by spatially aggregating the indices (pathway 8), or by aggregating the indicators separately and recalculating the indices at different spatial scales (pathway 9).
If perfectly commutative then the two approaches will yield the same result.
However, if the indicators that make up the indices use different area weights as part of their spatial aggregation (**Example 4**), then pathway 9 is the only one that gives the same index values across nested accounts.

@@

ECAs should be accompanied by detailed indicator documentation where a precise description of the steps in the aggregation process is presented.
Verbose descriptions can be supplemented or replaced by a standardised short-hand notation that we present in the following: We suggest using the terms as described in @fig-term and elsewhere in the paper: scale, truncate (abbreviated to *trunc.*), transform (*trans.*), spatial aggregation (*sp. agg.*), thematic aggregation (*th. agg.*) and spatial statistics (*sp. stat.*).
These terms can be abbreviated as shown in the parentheses, and placed together into a single string, where operations performed simultaneously are enclosed in square brackets.
For example, pathway 1 in @fig-pathways can be annotated as $scale - trunc. - trans. - sp. agg.$.
If the three normalisation steps are performed simultaneously this can be written as $[scale - trunc. - trans.] - sp. agg.$.
If there are more aggregation steps, then the spatial level at each step should be included in the description, e.g. for pathway 3: $sp. agg. (municipalities) - scale - sp. agg. (country)$.
"Municipality" and "country" could here be included as footnotes.
Note also that in the last example, normalisation only involved a scaling step.
The description of indicator pathways can be made even easier by referencing specific aggregation pathways by number, as we have done in this paper using @fig-pathways in this paper.
Methods used to change either the cover or the extent of the indicator data (see @tbl-terms) can also be included in the shorthand notation scheme.
The three operations we conceive as possible options are extrapolation (extr.), interpolation (inter.), and (sub)sampling.
We use the term spatial projection (projection for short) to encompass both extra- and interpolation.
Thus, pathway 7 in @fig-pathways can be written as $project - scale - sample - sp.agg. - trunc$.
The type of model used for the spatial projection can be added as a footnote.

@@

It may be premature, or not even possible, to prescribe an aggregation pathway to be used for all ECAs, or even a type of aggregation pathway to be used given the data at hand.
However, internal consistency within assessments should be possible, and this can make it easier to communicate and interpret the information conveyed through the indicator values.
This can be achieved by for example always scaling early (see recommendation number 2 above) and using the same aggregation levels (same spatial units) and same approach to step-wise aggregation across levels, for all indicators.

@@
